---
title: "Preprocessing rat data"
author: "Hannah Meyer"
date: "`r Sys.Date()`"
bibliography: referencesRat.bib
csl: plos-genetics.csl
output:    
    pdf_document:
    toc: true
toc_depth: 2
fig_caption: yes
vignette: >
    %\VignetteIndexEntry{Pre-pocessing of rat phenome dataset}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
devtools::load_all()
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r paths}
## directory ####
directory <- "~/data/LiMMBo/rat/rawdata"
```

# 1. Dataset
The dataset consists of an outbred population of rats descended from eight 
inbred progenitors. 2,006 outbred rats were phenotypes for 195 traits of 
biomedical relevance including hematological, immunological, cardiovascular and 
fear-related phenotypes. 1,407 of the rats were genotyped @Baud2013,@Baud2014.

The phenotype and genotype data was downloaded from [arrayexpress](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-2332/) and [figshare](http://dx.doi.org/10.6084/m9.figshare.943485). Supplementary 
information about the data (distribution, type, etc) was taken from @Baud2013,
@Baud2014.

```{r}
## from figshare ####
measures <- data.table::fread(paste(directory, "/figshare/measures.txt", sep=""), 
                  data.table=FALSE, stringsAsFactors = FALSE)

## from Baud et al (2014) ####
pheno_ScientificData <- data.table::fread(paste(directory, 
                                 "/PublicationResults/ScientificDataS1.csv", 
                                 sep=""),
                           data.table=FALSE, stringsAsFactors = FALSE, 
                           skip=2)[, -c(6:7)]
colnames(pheno_ScientificData) <- gsub(" ", "", colnames(pheno_ScientificData))
pheno_ScientificData$gsub <- gsub("_normalized_by_batch", "",
                                  pheno_ScientificData$Measure)
## from Baud et al (2013) ####
pheno_NatGenet <- data.table::fread(paste(directory, 
                              "/PublicationResults/", 
                              "SupplementaryTable1_Phenotypes.csv", sep=""),
                              data.table=FALSE, stringsAsFactors = FALSE, 
                        skip=3)
colnames(pheno_NatGenet) <- gsub(" ", "", colnames(pheno_NatGenet))
pheno_NatGenet$MeasureinGSCAN <- gsub("_bc", "", pheno_NatGenet$MeasureinGSCAN)
```

# 2. Sorting data
The rat phenotype and genotype will be used in a multi-variate association
analysis. The data downloaded from [figshare](http://dx.doi.org/10.6084/m9.figshare.943485)
contains both, phenotypes and measures used as covariates in downstream analysis.
In the following, `measures` is split based on their classification in the 
@Baud2014. Only phenotypes with approximately normal distribution will be kept
for this analysis, based on the original studies assumption of normality 
(as indicated with Mixed model in the analysis section of @Baud2014).

```{r sorting data}
covariate_names <- dplyr::filter(pheno_ScientificData, 
                                 Phenotypingtest == "Covariate")
phenotype_names <- dplyr::filter(pheno_ScientificData, 
                                 Phenotypingtest != "Covariate")

phenotype_names <- merge(phenotype_names, pheno_NatGenet[, -c(1:2,8)], 
                         by.x="gsub", by.y="MeasureinGSCAN", all=TRUE)
 
phenotype_names_notN <- dplyr::filter(phenotype_names, 
                                           Mappingmethod != "Mixed models",
                                           !is.na(Measure))
phenotype_names_N <- dplyr::filter(phenotype_names, 
                                  Mappingmethod == "Mixed models",
                                  !is.na(Measure))

phenotypes_normal <- measures[, colnames(measures) %in% 
                                  phenotype_names_N$Measure ]
rownames(phenotypes_normal) <- measures$SUBJECT.NAME

phenotypes_notNormal <- measures[, colnames(measures) %in% 
                                     phenotype_names_notN$Measure ]
rownames(phenotypes_notNormal) <- measures$SUBJECT.NAME


write.table(phenotypes_normal, paste(directory, "/phenotypes_normal.csv", 
                                     sep=""),
          row.names=TRUE, col.names=NA, quote=FALSE, sep=",")
write.table(phenotypes_notNormal, paste(directory, "/phenotypes_notNormal.csv", 
                                        sep=""),
            row.names=TRUE, col.names=NA, quote=FALSE, sep=",")

## covariates ####
covariates <- measures[, colnames(measures) %in% 
                           c("BW_week11", "BW_week14", "BW_week17", 
                             covariate_names$Measure) ]
covariates <- apply(covariates, 2, function(x) {
    if (! is.numeric(x)) return(as.numeric(as.factor(x)))
    if (is.numeric(x)) return(x)
})
rownames(covariates) <- measures$SUBJECT.NAME

# manually match covariate names from NatGenetc supp info to colnames of measure/
# covariates

uniqueCovariatesTested <- unique(unlist(phenotype2covs))
#[1] "sex"            "batch"          "is.albino"      "Haemalysis"     "BW_at_day28_pi" "BW_at_day9_pi"  "age"         
#[8] "BW_at_IPGTT"    "test_worked"
    
 
#colnames(covariates)
#[1] "BW_week11"              "age_dissection"         "analyser_ear_hole"      "batch"                 
#[5] "blood_pressure_time"    "cage"                   "date_IPGTT"             "date_of_birth"         
#[9] "died_of_EAE"            "sibship"                "Haemalysis_serum"       "is.albino"             
#[13] "litter_size"            "max_EAE_score"          "reliable_ear_hole"      "scanner_ear_hole"      
#[17] "EAE_score_at_sacrifice" "sex"                    "IPGTT_test_worked"      "BW_week14"             
#[21] "BW_week17"              "Exp_novel_cage"         "Exp_zero_maze"          "Exp_shuttlebox"        

## information about conversion of these variables directly from from Amelie Baud
colnames(covariates)[colnames(covariates) == "BW_week11"] <- "BW_at_IPGTT"
colnames(covariates)[colnames(covariates) == "BW_week14"] <- "BW_at_day9_pi"
colnames(covariates)[colnames(covariates) == "BW_week17"] <- "BW_at_day28_pi"
colnames(covariates)[colnames(covariates) ==  "age_dissection"] <- "age"
colnames(covariates)[colnames(covariates) == "Haemalysis_serum"] <- "Haemalysis"
colnames(covariates)[colnames(covariates) == "IPGTT_test_worked"] <- "test_worked"

relevantCovariates <- covariates[,colnames(covariates) %in% uniqueCovariatesTested]
write.table(relevantCovariates, paste(directory, "/covariates.csv", sep=""),
            row.names=TRUE, col.names=NA, quote=FALSE, sep=",")

```

# 3. Map covariates to phenotypes
For downstream analysis, map covariates to the phenotypes they were associated
with in the original study.
```{r covariate mapping}
phenotype_names_N <- phenotype_names_N[phenotype_names_N$Measure %in% 
                                           colnames(phenotypes_normal),]
phenotype_names_N <- phenotype_names_N[  match(colnames(phenotypes_normal), 
                                               phenotype_names_N$Measure),]

phenotype2covs <- lapply(1:nrow(phenotype_names_N), function(x) {
    unlist(strsplit(phenotype_names_N$Covariates[x], ","))
})
names(phenotype2covs) <- phenotype_names_N$Measure
```
saveRDS(phenotype2covs, paste(directory, "/phenotype2covs.rds", sep=""))