---
title: "Phenome imputation of a population of outbred rats"
author: "Hannah Meyer"
date: "`r Sys.Date()`"
bibliography: referencesRat.bib
csl: plos-genetics.csl
output:    
    pdf_document:
        toc: true
        toc_depth: 2
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Phenome imputation of a population of outbred rats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
devtools::load_all()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
The following vignette shows the imputation workflow for an additional dataset.
More detailed information about the analysis steps can be found in the vignette
'Phenome imputation of a panel of prototrophic haploid yeast segregants'. 

# 1. Dataset
The dataset consists of an outbred population of rats descended from eight 
inbred progenitors. 2,006 outbred rats were phenotypes for 195 traits of 
biomedical relevance including hematological, immunological, cardiovascular and 
fear-related phenotypes. 1,407 of the rats were genotyped @Baud2013,@Baud2014.

The phenotype and genotype data was downloaded from [arrayexpress](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-2332/) and [figshare](http://dx.doi.org/10.6084/m9.figshare.943485 (2014).).
The kinship was estimated from these genotypes: the *NrSamples x NrSNPs* 
genotypes $X$ were filtered for minor allele frequency of at least 5\% and the 
kinship estimated as $R=\frac{1}{NrSNPs}X^TX$.

```{r read data}
## directories ####
rawdir <- "~/data/LiMMBo/rat/rawdata/arrayexpress"
directory <- "~/data/LiMMBo/rat/processeddata"

## data ####
phenotypes_normal <- read.table(paste(directory, "/phenotypes_normal.csv", 
                                      sep=""), 
                            header=TRUE, row.names=1, stringsAsFactors=FALSE, 
                            sep=",")

covariates <- read.table(paste(directory, "/covariates.csv", sep=""), 
                            header=TRUE,  row.names=1, stringsAsFactors=FALSE, 
                            sep=",")

kinship <- read.table(paste(rawdir, "/HS_rats_kinship_norm.csv", sep=""), 
                            header=TRUE, stringsAsFactors=FALSE, 
                            sep=",")

common_samples <- colnames(kinship)[colnames(kinship) %in% 
                                        rownames(phenotypes_normal)]
kinship <- kinship[colnames(kinship) %in% common_samples,
                   colnames(kinship) %in% common_samples]
phenotypes_normal <- phenotypes_normal[rownames(phenotypes_normal) %in% 
                                           common_samples,]
phenotypes_normal <- phenotypes_normal[match(colnames(kinship), 
                                             rownames(phenotypes_normal)),]

covariates <- covariates[rownames(covariates) %in% 
                                           common_samples,]
covariates <- covariates[match(colnames(kinship), 
                                             rownames(covariates)),]

# for imputation purposes, combine covariate and phenotype data
combined_all <- cbind(covariates, phenotypes_normal)

## general parameters ####
cutoff <- 0.95
col <- c('#fc8d62','#8da0cb')
text_size <- 12
```


# 2. Pattern of missing data 
Aggregation plot (middle) where all existing combinations of missing (blue) and non-missing (orange) values in the traits are depicted. The bar chart on its right shows the frequencies of occurrence of the different combinations. The histogram on the top shows the frequency of missing values for each trait.

```{r missing data}
## a) distribution
plot_pattern_missingness(combined_all, directory=directory,
                         name="missing_data_pattern_allTraits")

frequency_missingness <- data.frame(missing=
                                        apply(combined_all, 2, function(x)
                                            length(which(is.na(x)))/length(x)))
frequency_missingness$complete <- 1 - frequency_missingness$missing

Traits2Keep <- frequency_missingness$missing <= 0.20
combined_filterTraits <- combined_all[, Traits2Keep]

per_sample_missingness <- data.frame(missing=
                                         apply(combined_filterTraits, 1, function(x) 
                                             length(which(is.na(x)))/length(x)))
per_sample_missingness$complete <- 1 - per_sample_missingness$missing

Samples2Keep <- per_sample_missingness$missing <= 0.20
kinship <- kinship[Samples2Keep, Samples2Keep]
combined <- combined_filterTraits[Samples2Keep,]
```

```{r initial phenotype filtering, message=FALSE}
plot_pattern_missingness(combined, directory=directory,
                         name="missing_data_pattern_filteredTraits")
```

Any sample with more than 20\% missing phenotype data is removed from further analyses, reducing the dataset to `r nrow(combined)` samples. The aggregation and frequency plots for this reduced dataset is shown above.


# 3. Missing data mechanism
Requirement for MAR data: missingness is predictable. If it can be demonstrated that one or more variables in the dataset are significantly correlated with missing values, missingness may be predictable
Visiually examining predictable missingness by correlating the observed values across all samples with each column of an indicator matrix, i.e. the missingness patterns per trait. If all values were observed for a given trait, all values in the indicator matrix in this column were equal to zero and the correlation between the trait and the missingness was set to NA. Overall, there is sufficient evidence for predictable missingness.

```{r missing data mechanism, fig.height=4, fig.width=4}
### MAR
corrMiss <- correlationMissingness(combined)
plot_correlation_missingness(data=corrMiss, savePdf=TRUE,
                             directory=directory, labelsize=0.3)
```


# 4. Dataset with no missing values
```{r fully phenoytped samples, fig.height=6, fig.width=6}
combined_noNA <- fullPhenotypes(combined)

## a) correlation between phenotypes 
corrPhenotypes <- correlationPhenotypes(combined)

plot_correlation_phenotypes(data_r=corrPhenotypes$r, 
                            data_p=corrPhenotypes$padjust, savePdf=TRUE,
                            directory=directory, labelsize=0.3)
```

Out of the  `r nrow(combined)` rats for which at least 80\% of phenotypes were measured, `r nrow(combined_noNA)` were fully phenotyped. The pairwise phenotype correlation across these rats is shown above.

# 5. Generate dataset with artificial missingness
Use the subset of the `r nrow(combined_noNA)` fully phenotyped rats and introducing missing values with a similar pattern of missingness as observed in the original dataset. The results for the real and simulated  dataset are similar in terms of frequencies and combinations of missing/non-missing traits.

```{r artifical missingness, message=FALSE}
combined_artificial <- artificialMissingness(data=combined, fulldata=combined_noNA,
                                     kinship=kinship, seed=3422)
plot_pattern_missingness(combined_artificial$data_addNA, directory=directory,
                         name="missing_data_pattern_simulated", savePdf=TRUE)

```

# 6. Impute artifically created missing data ####
The masked values in the dataset with artificial missingness were imputed with two genetic (phenix and mpmm) and two non-genetic (mvn and mice) methods. For imputation with mice, different predictor sets were analysed, based on the correlation of the measured phenotypes (All phenotypes as predictors, phenotypes with correlation > 0.1, 0.2 or greater 0.3). After imputation, the goodness of the imputation is evaluated by computing the Pearson correlation of the imputed values to the experimentally observed ones. Table \ref{tab:imputation} show those correlations across all traits for all imputation methods.

```{r impute artifical missingness, fig.height=3, fig.width=5, cache=TRUE,  fig.align='center', fig.show='hold', fig.cap = ""}
imputed <- imputeData(data=as.matrix(combined_artificial$data_addNA), 
                      fulldata=combined_noNA,
                      kinship=combined_artificial$kinship,
                      method=c("phenix",  "mvn", "mice"),
                      cutoff=cutoff, testing=TRUE)
knitr::kable(imputed$summary, 
caption="\\label{tab:imputation}Correlation of imputed vs real phenotype values")
plot_overview_correlation_imputation(imputed$cor, savePdf=TRUE,
                                     directory=directory,
                                     text_size=6)
```

```{r plot individual imputation results, fig.keep='all', fig.show='hold',fig.align='center', fig.height=3, fig.width=5, cache=TRUE, message=FALSE, warning=FALSE}

corr_plots_methods <- plot_individual_correlation_imputation(imputed$cor, 
                                                            savePdf=TRUE,
                                       directory=directory, text_size=6)
dummy <- lapply(corr_plots_methods, function(p) print(p))
```


# 7. Impute full data set
The traits that can be reliable imputed (correlation greater than 95\%) are selected and the missing values from the phenotype set with `r nrow(combined)` rats imputed.

```{r impute full data set, cache=TRUE}
imputable <-  data.frame(sapply(c("phenix", "mvn", "mice"), imputableTraits, 
                                  imputed, cutoff))

imputed_phenos <- imputeData(data=combined, methods= c("phenix", "mice", "mvn"), 
                       imputable=imputable, cutoff=0.95, testing=FALSE, 
                       kinship=as.matrix(kinship))
saveRDS(imputed_phenos, paste(directory, "/imputed_phenotypes.rds", sep=""))
```

```{r regress covariates from phenotypes}
phenotype2covs <- readRDS(paste(rawdir, "/phenotype2covs.rds", sep=""))

regress <- function(trait, pheno, cov, pheno2cov ) {
    which.cov <- which(colnames(cov) %in% pheno2cov[[trait]])
    summary(lm(pheno[,colnames(pheno) == trait] ~ cov[,which.cov]))$residuals
}

combined_phenix <- imputed_phenos$imp$phenix$imp
pheno_phenix <- combined_phenix[, which(colnames(combined_phenix) %in% colnames(phenotypes_normal))]
cov_phenix <- combined_phenix[, which(colnames(combined_phenix) %in% colnames(covariates))]
phenix_reg <- sapply(colnames(pheno_phenix), regress, pheno=pheno_phenix, 
                    cov=cov_phenix, pheno2cov=phenotype2covs)

write.table(phenix_reg, paste(directory, "/phenotypes_phenix_reg.csv", sep=""), 
            sep=",", col.names=NA, row.names=TRUE, quote=FALSE)

combined_mvn <- imputed_phenos$imp$mvn$imp
pheno_mvn <- combined_mvn[, which(colnames(combined_mvn) %in% colnames(phenotypes_normal))]
cov_mvn <- combined_mvn[, which(colnames(combined_mvn) %in% colnames(covariates))]
mvn_reg <- sapply(colnames(pheno_mvn), regress, pheno=pheno_mvn, 
                    cov=cov_mvn, pheno2cov=phenotype2covs)

write.table(mvn_reg, paste(directory, "/phenotypes_mvn_reg.csv", sep=""), 
            sep=",", col.names=NA, row.names=TRUE, quote=FALSE)


combined_mice<- imputed_phenos$imp$mice$imp
pheno_mice <- combined_mice[, which(colnames(combined_mice) %in% colnames(phenotypes_normal))]
cov_mice <- combined_mice[, which(colnames(combined_mice) %in% colnames(covariates))]
mice_reg <- sapply(colnames(pheno_mice), regress, pheno=pheno_mice, 
                    cov=cov_mice, pheno2cov=phenotype2covs)

write.table(mice_reg, paste(directory, "/phenotypes_mice_reg.csv", sep=""), 
            sep=",", col.names=NA, row.names=TRUE, quote=FALSE)

```

The pairwise correlation of the phenotypes imputed with mice, mvn and phenix is shown below.
In general, imputed phenotypes correlate strongly, while for two traits the correlation of the mice imputed traits
versus imputation with mvn and phenix falls below a correlation of 0.95. 

```{r corr imputed phenotypes}
commonImputable <- intersect(intersect(colnames(mice_reg), colnames(mvn_reg)), 
                             colnames(phenix_reg))

mice_mvn <- do.call(rbind, lapply(commonImputable, function(x) {
    tmp <- Hmisc::rcorr(mice_reg[,colnames(mice_reg) == x], 
                        mvn_reg[,colnames(mvn_reg) == x])
    return(data.frame(p=tmp$P[1,2], r2=tmp$r[1,2], comparison="mice_mvn", trait=x))
}))

mice_phenix <- do.call(rbind, lapply(commonImputable, function(x) {
    tmp <- Hmisc::rcorr(mice_reg[,colnames(mice_reg) == x], 
                        phenix_reg[,colnames(phenix_reg) == x])
    return(data.frame(p=tmp$P[1,2], r2=tmp$r[1,2], comparison="mice_phenix", trait=x))
}))

phenix_mvn <- do.call(rbind, lapply(commonImputable, function(x) {
    tmp <- Hmisc::rcorr(phenix_reg[,colnames(phenix_reg) == x], 
                        mvn_reg[,colnames(mvn_reg) == x])
    return(data.frame(p=tmp$P[1,2], r2=tmp$r[1,2],comparison="mvn_phenix", trait=x))
}))

compare_corr <- rbind(mice_mvn, mice_phenix, phenix_mvn)
compare_corr$comparison <- as.factor(compare_corr$comparison)

p <- ggplot(data=compare_corr, aes(x=comparison, y=r2, 
                                   color=comparison))
p + geom_boxplot(outlier.colour = NA) + 
    geom_label(data=dplyr::filter(compare_corr, r2 < 0.95), aes(y=r2, 
                                                                x=comparison,
                                                                label=trait),
               nudge_y = -0.002, size=3) +
    geom_jitter(width = 0.2) +
    scale_color_manual(values=c('#1b9e77','#d95f02','#7570b3'), guide=FALSE) +
    ylab("Pearson Correlation") +
    theme_bw()

```

\newpage
# 8. References